---
title: "**S2208 MATH8050 Data Analysis - Section 001: Homework 2 Due on 09/21/22**"
author: "**Adithya Ravi, C09059838**"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=TRUE, include=TRUE, results="hide"}
# load packages 


sessionInfo()
```

# Solutions

## 1

1. Write an R function called `f()` that can take a numeric vector as input argument to implement this function using `if` (or `ifelse`) statements, and plot the function in the interval $[-2, 2]$.


```{r}
f <- function(x){
  if(x <= 0) {
    -(x^3)
  }
  else if(x > 0 & x <= 1){
    x^2
  }
  else{
    sqrt(x)
  }
}

f <- Vectorize(f, vectorize.args = 'x')
curve(f(x), from = -2, to = 2)
```

## 2

$$ 
p(x|\theta) = \text{Poisson}(x|\theta) = \exp(-\theta) \frac{\theta^x}{ x!},
$$ for $x \in \{0,1,2,\ldots \}$ (and is 0 otherwise). Suppose $X_1,\ldots,X_n \stackrel{iid}{\sim} \text{Poisson}(\theta)$ given $\theta.$ Please work on the following problems. 

a.  Find the likelihood function and the maximum likelihood estimator of $\theta$. 

The  n observations are independent. As a consequence, the likelihood function is equal to the product of their probability mass functions:
&nbsp;
$$\Pi_{i=1}^{n}\frac{\theta^{x_i}e^{-\theta}}{x_i!}$$
&nbsp;
It will be easier to find the value of $\theta$ that maximizes this quantity if we take the log:
$$\log (\Pi_{i=1}^{n}\frac{\theta^{x_i}e^{-\theta}}{x_i!}) = \sum_{i=1}^{n}\log(\frac{\theta^{x_i}e^{-\theta}}{x_i!}) = \sum_{i=1}^n x_i\log(\theta) - \sum_{i=1}^n \theta - \sum_{i=1}^n \log (x_i!)$$
$$=\log(\theta)\sum_{i=1}^nx_i - n\theta - \sum_{i=1}^n \log (x_i!)$$ 
&nbsp;
To find the value of $\theta$ that maximizes this equation, we take the derivative, set the derivative equal to zero, and solve for $\theta$:
$$0 = \frac{1}{\theta}\sum_{i=1}^nx_i - n$$
$$n\theta = \sum_{i=1}^n x_i$$
$$\theta = \frac{\sum_{i=1}^n x_i}{n} = \bar{x}$$"
&nbsp;
Therefore, the estimator is just the sample mean of the n observations in the sample.
&nbsp;

b. Show that the $\hat{\theta}$ is the unique maximum likelihood estimator (MLE). 

It is indeed unique, because the negative log-likelihood is strictly convex.

&nbsp;
&nbsp;

c. Further assume that the observations are $\{0, 2, 5, 6, 10, 15, 8, 7\}$. Plot the likelihood function as a function of $\theta$, and maximize the likelihood function numerically using the `optim` function from the base package. Denote the numerical solution to the MLE by $\hat{\theta}_1$. Is the numerical solution $\hat{\theta}_1$ the same as the true MLE $\hat{\theta}=\bar{x}=6.625$? 




&nbsp;
&nbsp;

d. Assume that we want to put a prior distribution on $\theta$ and perform Bayesian estimation. Your prior on $\theta$ is a gamma distribution 
\begin{align*}
p(\theta) = \text{Ga}(\theta|a, b) = \frac{b^a}{\Gamma(a)} \theta^{a-1} \exp(-b\theta) \I(\theta>0). 
\end{align*}
Derive the posterior distribution of $\theta$. 


&nbsp;
&nbsp;

e. Consider two sets of values for $a, b$: (1) $a=0.1, b=0.1$, which corresponds to a non-informative prior, and (2) $a=2, b=5$, which corresponds to an informative prior. Plot the prior, likelihood, and posterior, and arrange the plots from (1) and (2) in two panels in a row.



&nbsp;
&nbsp;

f. Derive the expression of the marginal likelihood with the $\text{Ga}(a, b)$ prior on $\theta$.



&nbsp;
&nbsp;

g. Derive the expression of the posterior predictive distribution with the $\text{Ga}(a, b)$ prior on $\theta$.

&nbsp;
&nbsp;
&nbsp;

## 3

a. Show that the hat matrix is symmetric and idempotent: i.e., $H^\top = H$ and  $HH=H$.

To show that the hat matrix is symmetric we have to prove that  $H^\top = H$
We know that 
$$H = X(X'X)^{-1}X'$$
The transpose of a product is the product of the transposes in reverse order. Therefore,
$$ H' = [X(X'X)^{-1}X']' = X [(X'X)^{-1}]' X'$$
We know that $(X'X)$ is symmetric and therefore the inverse $(X'X)^{-1}$ must also be symmetric. If $(X'X)^{-1}$ is symmetric, then by definition:
$$[(X'X)^{-1}]' = (X'X)^{-1}$$
Substituting this in our previous equation we get 
$$H' = [X(X'X)^{-1}X']' = X [(X'X)^{-1}]' X' = X(X'X)^{-1}X' = H$$
Thus, $ H' = H $
&nbsp;
&nbsp;
To show that H is idempotent we need to prove that  $HH=H$
We know that $H = X(X'X)^{-1}X'$. Thus,
$$HH = [X(X'X)^{-1}X'] [X(X'X)^{-1}X']$$
$$ HH =  X (X'X)^{-1} (X'X) (X'X)^{-1} X'$$
Since $ (X'X)^{-1} (X'X) = I$ (Identity matrix), we can rewrite this as,
$$  HH =  X (I )(X'X)^{-1} X'$$
$$  HH =  X (X'X)^{-1} X' = H$$
Thus, $HH=H$

b. Show that the residual vector $\mathbf{e} = \mathbf{y} - \hat{\mathbf{y}}$ has mean zero, i.e., $E(\mathbf{e}) = 0$ and variance $Var(\mathbf{e}) = \sigma^2 (I - H)$. 

The residual vector can also be written as:
$$\mathbf{e} = \mathbf{y} - \hat{\mathbf{y}} = \mathbf{y} - \mathbf{X}\mathbf{b} $$
Therefore,
$$ \mathbf{y} = \mathbf{X}\mathbf{b} +  \mathbf{e}$$
Recall the normal equation : 
$$(X'X)\hat{b} = X'y$$
Now substituting $ y = X \hat{b} + e $ in the above equation we get,
$$ (X'X)\hat{b} = X'(X\hat{b}+e) \\
(X'X)\hat{b} = (X'X)\hat{b} + X'e \\
X'e = 0 $$
From $X'e = 0$ we can derive that if there is a constant the first column in X i.e, $X_1 $ will be a column of ones. This means that for the first element in the $X′e$ vector (i.e, $X_{11} * e_1 + X_{12} * e_2... + X_{1n} * e_n$) to be zero,
$$\sum{e_i} = 0$$.
If $\sum{e_i} = 0$, we can derive that the mean of the residual vector is also zero:
$$ E[e] = \frac{\sum{e_i}}{n} = 0$$
&nbsp;
To show the variance of residuals we start with $e = Y-\hat{Y}$. Since $\hat{Y} = HY$,
$$ e = Y - HY = (I - H)Y \\
\sigma^2[e] = (I-H)\sigma^2[I-H]'
$$
But we can see that:
$$ \sigma^2[Y] = \sigma^2[e] = \sigma^2[I] $$
This means that:
$$ \sigma^2[e] = \sigma^2(I - H)I(I - H) \\
\sigma^2[e] = \sigma^2(I - H)(I-H)
$$
Since $(I-H)$ is idempotent we have, 
$$ \sigma^2[e] = \sigma^2[I-H]$$
&nbsp;

c. Show that the MSE is an unbiased estimator of $\sigma^2$, i.e., $E(MSE) = \sigma^2$. 

To prove that $E(MSE) = \sigma^2$ , we have to prove that $E(MSE) = (n-2)\sigma^2$.
Under the usual notation,
$$ Y = Xb+e \\
\hat{Y} = X \hat{b} \\
\hat{b} = (X'X)^{-1}X'Y \\
   \hat{b'} = Y'X(X'X)^{-1}
$$
&nbsp;
Now,
$$
\sum{(Y_i-\hat{Y_i})}^2 = (Y_i - \hat{Y_i})'(Y_i - \hat{Y_i}) \\ 
=(X(b-\hat{b})+e)'(X(b-\hat{b})+e) \\
= (b-\hat{b})'X'X(b-\hat{b})+e'X(b-\hat{b}) + (b-\hat{b})'X'e+e'e
$$
&nbsp;
Simplifying the individual terms:
TERM 1
$$ (b-\hat{b})'X'X(b-\hat{b}) = (b- (X'X)^{-1}X'Y)'X'X(b-(X'X)^{-1}X'Y \\
= (b'-Y'X(X'X)^{-1})X'X(b-(X'X)^{-1}X'Y) \\
= b'X'Xb-Y'Xb-b'(X'X)^{-1}X'Y+Y'X(X'X)^{-1}X'Y \\
= b'X'Xb-(b'X'+e')Xb-b'(X'X)(X'X)^{-1}X'Y + (b'X'+e')X(X'X)^{-1}X'Y (substituting) \\ 
= e'Xb+e'X(X'X)^{-1}X'Y (cancelled) \\
= -e'Xb+e'X(X'X)^{-1}X'(Xb+e) (substituting) \\ 
=e'X(X'X)^{-1}X'e

$$
&nbsp;
TERM 2
$$
e'X(B-\hat{b}) \\ 
= e'X(b-(X'X)^{-1}X'Y) \\
= e'X[b-(X'X)^{-1}X'(Xb+e)] \\
= e'X [b-(X'X)^{-1}X'Xb-(X'X)^{-1}X'e] \\
= -e'X(X'X)^{-1}X'e
$$
&nbsp;
TERM 3 is a transpose of TERM 2
Now substituting in we get, 
$$
\sum{(Y_I-\hat{Y_i})^2} = e'X(X'X)^{-1}X'e - e'X(X'X)^{-1}X'e - e'X(X'X)^{-1}X'e+e'e \\ 
= e'e-e'X(X'X)^{-1}X'e= e'e-e'Pe
$$
 P is the projection matrix which is symmetric and idempotent
Now we calculate the expectation:
$$
E[\sum{(Y_i-\hat{Y_i})^2}] = E(e'e - e'Pe) = E(e'e)-E(e'Pe)= n\sigma^2 - \sigma^2(2P)
$$
4. (14pts total, equally weighted) Refer to the Lecture 4. For the normal simple linear regression model
$$ y_i | x_i, \boldsymbol \beta, \sigma^2 \stackrel{ind}{\sim} N(\beta_0 + \beta_1 x_i, \sigma^2)$$ with $i=1, \ldots, n$.

a. Show that 

$$
\begin{aligned}
Var(\widehat{\boldsymbol{\beta}}) & =  \sigma^2 (\mathbf{X}^\top \mathbf{X})^{-1} =
\left[ \begin{matrix}
\sigma^2 \left(\frac{1}{n}+\frac{\overline{x}^2}{ \sum_{i=1}^n(x_i -\overline{x})^2}\right) &
\frac{-\sigma^2 \overline{x}}{\sum_{i=1}^n(x_i - \overline{x})^2}\\
\frac{-\sigma^2 \overline{x}}{\sum_{i=1}^n(x_i - \overline{x})^2} &
\frac{\sigma^2}{\sum_{i=1}^n(x_i -\overline{x})^2} \end{matrix} \right],
\end{aligned}
$$
&nbsp;
&nbsp;

b. Assume that $\sigma^2$ is given. Let $\hat{y}_h:=\hat{\beta}_0 + \hat{\beta}_1 x_h$ denote the estimator of $y_h$ at new covariate $x_h$. Show that the posterior predictive distribution of the new response $y_h$ at new covariate $x_h$ is 
$$y_h \mid \mathbf{y}, \mathbf{X}, x_h, \sigma^2 \sim \mathcal{N}\left(\hat{y}_h, \sigma^2\left\{1 + \frac{1}{n} + \frac{(x_h - \bar{x})^2}{ \sum_{i=1}^n(x_i - \bar{x})^2} \right\} \right).$$
